%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[
% twocolumn,
% hf,
]{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
%% auto break lines
\lstset{breaklines=true}


%% Custom Listing style by Pascal
\usepackage{xcolor}
\usepackage{tcolorbox}

\definecolor{lightgray}{gray}{0.95}

\lstdefinestyle{promptstyle}{
    backgroundcolor=\color{lightgray},
    basicstyle=\ttfamily\small,
    frame=single,
    rulecolor=\color{black},
    breaklines=true,
    columns=fullflexible,
    captionpos=b,
    showstringspaces=false,
    keepspaces=true,
    xleftmargin=0pt,
    xrightmargin=0pt,
    aboveskip=1em,
    belowskip=1em
}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2025}
\copyrightclause{Copyright for this paper by its authors.
  Use permitted under Creative Commons License Attribution 4.0
  International (CC BY 4.0).}


%%
%% This command is for the conference information
\conference{CLEF 2025 Working Notes, 9 -- 12 September 2025, Madrid, Spain}

%%
%% The "title" command
\title{Deep Retrieval at CheckThat! 2025: Identifying Scientific Papers from Implicit Social Media Mentions via Hybrid Retrieval and Re-Ranking}
\title[mode=sub]{Notebook for the CheckThat! Lab at CLEF 2025}

%\tnotemark[1]
%\tnotetext[1]{You can use this document as the template for preparing your
%  publication. We recommend using the latest version of the ceurart style.}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1,2,3]{Pascal J. Sager}[%
orcid=0000-0002-8084-2317,
email=pascaljosef.sager@uzh.ch,
]
\cormark[1]

\author[1]{Ashwini Kamaraj}[%
email=ashwini.kamaraj@uzh.ch,
]

\author[3]{Benjamin F. Grewe}[%
orcid=0000-0001-8560-2120,
email=bgrewe@ethz.ch,
]

\author[2,4]{Thilo Stadelmann}[%
orcid=0000-0002-3784-0420,
email=stdm@zhaw.ch,
]

\address[1]{University of Zurich, Rämistrasse 71, 8006 Zurich, Switzerland}
\address[2]{Centre for Artificial Intelligence, Zurich University of Applied Sciences, Technikumstrasse 71, 8401 Winterthur, Switzerland}
\address[3]{Institute of Neuroinformatics, ETH Zurich and University of Zurich, Winterthurerstrasse 190, 8057 Zurich, Switzerland}
\address[4]{European Centre for Living Technology, Dorsoduro 3246, 30123 Venice, Italy}


%% Footnotes
\cortext[1]{Corresponding author.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\input{sections/01_abstract}


%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
    Information retrieval \sep
    Scientific document search \sep
    Social media fact verification \sep
    Fact-checking \sep
    CLEF
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
\input{sections/02_fig_motivation_clef}
\input{sections/02_introduction}

\section{Related Work}
\input{sections/03_related_work}

\section{Methodology}
\input{sections/04_fig_overview_clef}
\input{sections/04_methods}

\section{Results} \label{sec:results}
\input{sections/05_results}


\section{Discussion and Conclusion}
\input{sections/06_discussion}

%%
%% The acknowledgments section is defined using the "acknowledgments" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acknowledgments}
  We thank Prof. Dr. Simon Clematide and Andrianos Michail for guiding the research, engineering, and writing process.
  We also thank the Department of Computational Linguistics at the University of Zurich and the Centre for Artificial Intelligence at the Zurich University of Applied Sciences for providing computational resources.
\end{acknowledgments}

%% The declaration on generative AI comes in effect
%% in Janary 2025. See also
%% https://ceur-ws.org/GenAI/Policy.html
\section*{Declaration on Generative AI}
\noindent During the creation of this work, the authors used ChatGPT\footnote{\url{https://chatgpt.com}} to refine the pre-written text. Further, the authors used Grammarly\footnote{\url{https://www.grammarly.com}} for spell checking.
 After using these tools, the authors reviewed and edited the content as needed and take full responsibility for the publication’s content. 

%%
%% Define the bibliography file to be used
\bibliography{sources_cleaned}

%%
%% If your work has an appendix, this is the place to put it.
\appendix
\input{sections/10_appendix}


\end{document}

