%
In this work, we demonstrate how bidirectional and autoregressive objective functions influence the structure of the query-key matrix $\bm{W}_{qk}$ in self-attention, enhancing our understanding of Transformer models.
%
Our mathematical framework shows that bidirectional training induces symmetric structures in $\bm{W}_{qk}$, whereas autoregressive training results in matrices characterized by directionality and column dominance.
%
To empirically validate our analysis, we develop and apply symmetry and directionality scores to various Transformer encoder and decoder models across multiple modalities, including text, audio, and images. 
%
Our results reveal that bidirectionally trained encoder models exhibit high symmetry, while autoregressively trained decoder models demonstrate strong directionality, thereby supporting the predictions of our mathematical framework.
%
%To validate our analysis empirically we define a symmetry and directionality score, showing consistent results across different Transformer architectures, input modalities, and layers within a model.
%
This suggests that self-attention inherently reflects these structural properties, contributing to the mechanistic interpretability of Transformer models.
%
%Our results thus indicate fundamental characteristics of the self-attention mechanism, contributing to a deeper mechanistic interpretability of Transformer models.
%
Finally, we leverage our findings to improve convergence speed during bidirectional training by initializing $\bm{W}_q$ and  $\bm{W}_k$ matrices such that $\bm{W}_{qk}$ is symmetric.
%
%This result shows a practical consequence of our findings beyond their implications for interpretability.
%
While our findings mark an initial step toward leveraging symmetry for more efficient Transformer training, further research is required to assess the scalability of symmetric initialization in large-scale models and across diverse domains.
%
% Future research should validate our initialization strategy across diverse models and test its scalability in larger architectures.
%
Furthermore, it is important to explore strategies for leveraging the directionality structures of decoder-only models.
%
For instance, incorporating structural constraints into the objective function or weight regularization could enhance training efficiency and stability in autoregressive settings.
%

Intriguingly, we found that there are similarities between the structure of the attention weights that we find and existing connectivity patterns in cortex, as areas dedicated and image comprehension show symmetric connectivity patterns  \cite{song2005highly,wang2006heterogeneity}, while areas dedicated to language are highly directional \cite{peng2024directed}. Further studies could use those patterns to connect self-attention mechanisms and models of information processing in the brain \cite{gershman2025key}.

By bridging theoretical insights with practical improvements, our work not only advances the interpretability of self-attention but also provides a foundation for optimizing Transformer architectures.
%
Ultimately, these findings contribute to a deeper understanding of the mechanisms governing self-attention, paving the way for more reliable and efficient Transformer-based models.