% =============================================================================
\chapter{Agents for Computer Use: A Comprehensive Survey}
\label{chap:acu}
% =============================================================================

% ─────────────────────────────────────────────────────────────────────────────
% PUBLICATION HEADER
% ─────────────────────────────────────────────────────────────────────────────

\begin{tcolorbox}[
    colback=gray!6,
    colframe=gray!35,
    arc=3pt,
    boxrule=0.4pt,
    left=8pt, right=8pt, top=6pt, bottom=6pt
]
\small
\textbf{Publication basis of this chapter}

\medskip
Sager, P.\,J., Meyer, B., Yan, P., von Wartburg-Kottler, R., Etaiwi, L.,
Enayati, A., Nobel, G., Abdulkadir, A., Grewe, B.\,F., \& Stadelmann, T.
(2026). A comprehensive survey of agents for computer use: Foundations,
challenges, and future directions. \textit{Journal of Artificial Intelligence
Research}, \textit{85}, 34. \cite{acu_survey}

\medskip
\textbf{Status:} Published.
\end{tcolorbox}

\bigskip

% ─────────────────────────────────────────────────────────────────────────────
\section*{Contribution Statement}
\label{sec:acu_contribution}
\addcontentsline{toc}{section}{Contribution Statement}
% ─────────────────────────────────────────────────────────────────────────────

The thesis author holds shared first authorship on this work with Benjamin
Meyer. The nature of the collaboration was deeply iterative and the
contributions of both first authors are interleaved throughout the paper in a
way that makes clean attribution of individual sections difficult. The following
account reflects the thesis author's best reconstruction of the intellectual
trajectory.

The project originated with the thesis author, who identified the absence of
a technology-agnostic, theoretically grounded taxonomy as a gap in the existing
ACU literature and initiated the survey independently. The initial structure of
the taxonomy and the selection of the three analytical perspectives — domain,
interaction, and agent — were proposed in a first draft by the thesis author.
Benjamin Meyer subsequently refined the taxonomy structure substantially,
improving its internal consistency and the precision of its category boundaries,
after which the thesis author revised the framework again. The final taxonomy
reflects this multi-pass collaborative development and cannot be attributed to
either first author alone.

The paper collection and classification process was carried out collectively by
all authors: each author independently read and classified a subset of the
surveyed papers into a shared structured spreadsheet, using the evolving
taxonomy as a classification schema. The survey structure, including the
identification and articulation of the six research gaps, was derived
inductively from the classification data through discussions involving both
first authors and with input from the senior authors Ahmed Abdulkadir,
Benjamin Grewe, and Thilo Stadelmann.

The framing of the survey's findings within the broader thesis argument
presented in this chapter — specifically, the interpretation of the six
research gaps as symptoms of structural limitations in the learning paradigm
rather than as incidental engineering shortcomings — is the intellectual
contribution of the thesis author and was developed in the context of this
thesis rather than the original paper.


% ─────────────────────────────────────────────────────────────────────────────
\section{Chapter Introduction}
\label{sec:acu_intro_thesis}
% ─────────────────────────────────────────────────────────────────────────────

The argument developed in Chapter~\ref{chap:introduction} rests on a specific
empirical claim: that the failures of current deep learning systems are not
distributed randomly across applications but form a consistent pattern,
one that is predictable from the structural properties of what
gradient-based learning produces. Establishing this claim requires an
empirical anchor — a domain in which the failure modes can be observed
concretely, measured systematically, and connected to the underlying learning
paradigm. Agents for computer use (ACUs) provide precisely this anchor.

ACUs are systems that execute complex tasks on digital devices given
instructions in natural language, automating workflows by controlling software
through low-level actions such as mouse clicks and touchscreen gestures. The
combination of natural language understanding, visual perception, multi-step
planning, and real-time interaction makes ACUs one of the most demanding
evaluation settings for contemporary AI. They are also among the most
practically consequential: a reliable general-purpose ACU would constitute a
qualitative shift in how AI systems are deployed. The gap between this
potential and the current reality, documented in detail in this chapter, is
the gap that motivates the rest of this thesis.

The contribution of this chapter is a comprehensive survey and taxonomy of the
ACU literature, covering 87 original research papers and 33 datasets published
between 2018 and 2024. We structure the literature using a three-perspective
taxonomy grounded in established agent theory — domain, interaction, and agent
perspectives — and apply it to identify six persistent research gaps. These
gaps are not presented here merely as a field-specific finding; they serve as
the empirical foundation for the structural diagnosis developed in
Part~\ref{part:diagnosis}.

Section~\ref{sec:acu_methodology} describes the survey methodology.
Sections~\ref{sec:field} through~\ref{sec:evaluation_agent} constitute the
core survey content, reproduced from the published paper with minor
reformatting for thesis consistency. Section~\ref{sec:acu_chapter_summary}
provides a new chapter summary that draws out the specific implications of the
survey findings for the thesis argument.

A note on scope: the broader literature on ACUs and the prior surveys in this
space are reviewed in Chapter~\ref{chap:relatedwork}, Section~\ref{sec:rw_agents}.
The methodology section below discusses methodological prior work specifically
relevant to the survey design; general contextualisation of the ACU field in
the literature is not repeated here.


% ─────────────────────────────────────────────────────────────────────────────
%
%   ██████╗ ██████╗ ██████╗ ███████╗    ██████╗  ██████╗ ██████╗ ██╗   ██╗
%  ██╔════╝██╔═══██╗██╔══██╗██╔════╝    ██╔══██╗██╔═══██╗██╔══██╗╚██╗ ██╔╝
%  ██║     ██║   ██║██████╔╝█████╗      ██████╔╝██║   ██║██║  ██║ ╚████╔╝
%  ██║     ██║   ██║██╔══██╗██╔══╝      ██╔══██╗██║   ██║██║  ██║  ╚██╔╝
%  ╚██████╗╚██████╔╝██║  ██║███████╗    ██████╔╝╚██████╔╝██████╔╝   ██║
%   ╚═════╝ ╚═════╝ ╚═╝  ╚═╝╚══════╝   ╚═════╝  ╚═════╝ ╚═════╝    ╚═╝
%
%   ═══════════════════════════════════════════════════════════════════════
%   INSERT SECTIONS 2–9 OF THE PUBLISHED PAPER HERE
%   (from \section{Formalizing the Problem} / \label{sec:field}
%    through \section{Evaluation} / \label{sec:evaluation_agent})
%
%   Renumber all figures and tables to the thesis scheme:
%     Figure 1  →  Figure~\ref{fig:acu_example}   (prefix: acu_)
%     Table 1   →  Table~\ref{tab:acu_...}         (prefix: acu_)
%
%   Replace all \cref{sec:X} with \ref{sec:X} if using cleveref,
%   or adjust cross-reference style to match thesis preamble.
%
%   The Appendix material from the paper (Appendix A1–A3 and the
%   supplementary tables) should be placed in the thesis Appendix
%   chapter, referenced here as Appendix~\ref{app:acu}.
%   ═══════════════════════════════════════════════════════════════════════
%
% ─────────────────────────────────────────────────────────────────────────────


\input{imported_papers/ACU_JAIR/sections/02_field_of_agents}
\input{imported_papers/ACU_JAIR/sections/03_domain_perspective}
\input{imported_papers/ACU_JAIR/sections/04_interaction_perspective}
\input{imported_papers/ACU_JAIR/sections/04_1_observation_space}
\input{imported_papers/ACU_JAIR/sections/04_2_action_space}
\input{imported_papers/ACU_JAIR/sections/05_agent_perspective}
\input{imported_papers/ACU_JAIR/sections/06_datasets}
\input{imported_papers/ACU_JAIR/sections/07_evaluation}
\input{imported_papers/ACU_JAIR/sections/08_challenges}


% ─────────────────────────────────────────────────────────────────────────────
\section{Chapter Summary}
\label{sec:acu_chapter_summary}
% ─────────────────────────────────────────────────────────────────────────────

This chapter has presented a comprehensive survey and taxonomy of agents for
computer use, covering 87 original research papers and 33 benchmark datasets
spanning the period 2018 to 2024. The three-perspective taxonomy — domain,
interaction, and agent — provides a technology-agnostic organisational
framework that bridges the previously fragmented literatures on
reinforcement-learning-based agents and foundation-model-based agents, enabling
systematic comparison across architectures and training paradigms.

Three structural trends characterise the current state of the field. The field
is transitioning from specialised, task-specific agents toward foundation-model-based
agents that leverage pre-trained vision-language models for open-ended reasoning.
The dominant observation modality is shifting from structured text
representations, such as HTML and accessibility trees, toward raw screenshots,
reflecting both the fragility of text-based observations and the improving
visual capabilities of foundation models. Behaviour cloning from human
demonstrations has emerged as the dominant learning paradigm among agents that
do learn from experience, though its data requirements and limited adaptability
represent significant practical constraints.

The primary finding of the survey, and its central contribution to the
thesis argument, is the identification of six persistent research gaps that
recur across the surveyed systems independently of their specific architecture,
training procedure, or platform:

\begin{enumerate}
    \item \textbf{Insufficient generalisation.} ACU systems fail reliably when
    interface layouts, task formulations, or environmental conditions differ
    from those encountered during training. The failure pattern is not one of
    missing data but of missing representational structure: systems cannot
    recombine familiar components in novel configurations.

    \item \textbf{Inefficient learning.} Current systems require large volumes
    of demonstration data and do not transfer learned behaviour effectively to
    related tasks. Sample inefficiency is systematic, not incidental, and
    reflects the absence of structural priors that would make infrequently
    encountered situations tractable.

    \item \textbf{Limited planning.} Reliable execution of long-horizon tasks
    requires maintaining a coherent model of task state and anticipating the
    effects of future actions. Current systems fail along both dimensions, either
    drifting from the goal over long autoregressive sequences or relying on world
    models whose predictions diverge exponentially with rollout length.

    \item \textbf{Low task complexity in benchmarks.} Widely used benchmarks
    evaluate agents on tasks that are systematically simpler than real-world
    computer use in sequence length, environmental variability, and success
    criterion complexity. Strong benchmark performance does not reliably predict
    strong deployment performance.

    \item \textbf{Non-standardised evaluation.} The absence of shared evaluation
    protocols makes direct comparison between systems unreliable. Reported
    success rates are not comparable across works, inflating the apparent
    progress of the field.

    \item \textbf{Disconnect between research and practical conditions.} ACU
    systems are developed and evaluated in static, deterministic, idealised
    environments. Real-world deployment introduces dynamic interfaces,
    non-stationarity from software updates, privacy constraints, and the need
    for conditional rather than full autonomy — properties that are absent from
    virtually all existing research settings.
\end{enumerate}

These six gaps are not independent failures of individual systems. They are
manifestations of a common pattern: current AI systems exhibit high competence
within the statistical regime of their training distribution and systematic
brittleness outside it. The specificity of this pattern demands a specific
explanation.

The standard response in the field is to attribute the gaps to insufficient
scale or to benchmark limitations, and to predict that they will close as
models and datasets grow. The evidence presented in this chapter does not
support this prediction. The generalisation and planning failures persist in
the largest and most capable foundation models currently evaluated on ACU
benchmarks. The benchmark complexity gap is not narrowing: as benchmarks
become more realistic, the gap between evaluated performance and the minimum
required for practical deployment remains substantial. And the disconnect
between research conditions and deployment conditions is structural — it
reflects assumptions baked into experimental design, not a shortfall in
model capacity that additional compute can correct.

Part~\ref{part:diagnosis} takes up the question of what these failures do
indicate. Chapter~\ref{chap:selfattention} establishes, at the level of
mathematical proof, that the representational geometry produced by
gradient-based training on standard objectives is determined by the structure
of the training signal rather than by the causal structure of the world.
The six gaps identified here are, under this account, predictable consequences
of a learning paradigm that encodes statistical correlations rather than
invariant structure. This reframing has direct implications for what kinds of
interventions can address the gaps — a question taken up in Parts~\ref{part:constraints}
and~\ref{part:paradigm}.