%
\input{ICML2025-rebuttal/tables/supp-table-vit-configs}
%
In parallel to the experiments described in Section~\ref{sec-symmetric-initialization}, where symmetric initialization of the query-key weight matrix $\bm{W}_{qk}$ is investigated for BERT models, we extend the investigation to the vision domain. Specifically, we evaluate the impact of symmetric initialization in vision transformers (ViTs) \citep{dosovitskiyImageWorth16x162021}, training a 6-layer ViT on CIFAR-10 \citep{krizhevsky2009learning} and a 12-layer ViT on ImageNet-1k \citep{deng_imagenet_2009}, using the same initialization strategy as applied to BERT models (see Appendix~\ref{sec:exp_bertmodels_models} for implementation details). The architectural configurations for both models are summarized in Table~\ref{tab:vit_configs}.

Training is performed using the AdamW optimizer \citep{Loshchilov_Hutter_2019} with weight decay and a cosine annealing learning rate schedule. A linear warm-up phase is used during the first $30$ epochs, with an initial learning rate scaled by a factor of $0.033$. The ImageNet-1k model is trained for $200$ epochs using a per-device batch size of $256$ and gradient accumulation over two steps, while the CIFAR-10 model is trained for $500$ epochs with a batch size of $128$ and no gradient accumulation. Both setups use a base learning rate of $0.003$, optimizer hyperparameters $\beta_1 = 0.9$ and $\beta_2 = 0.999$, and automatic mixed-precision training (fp16). We apply advanced data augmentation tailored to each dataset, including mixup ($\alpha = 0.2$) \citep{zhang2018mixup}, cutmix ($\alpha = 1.0$) \citep{Yun_cutmix_2019}, and label smoothing ($0.11$ for ImageNet-1k and $0.01$ for CIFAR-10). An exponential moving average (EMA) of the model weights \citep{Tarvainen_mean_2017} is maintained throughout training (decay rate = $0.99998$, update frequency = $32$ steps).
%
\input{ICML2025-rebuttal/tables/sup-table-results-symmetric-initialization-vit}

Contrary to the trends observed in our language model experiments, symmetric initialization does not result in faster convergence or improved final performance for the vision transformer models. As shown in Table~\ref{table:vit-symm-results}, training and evaluation losses, as well as top-1 accuracy, remained nearly identical between the standard and symmetric initialization conditions across both datasets.
Importantly, the results with symmetric initialization are not degraded, indicating that such initialization is at least performance-neutral in our vision setting. While our findings suggest that symmetric initialization does not yield immediate benefits for ViTs under the given training regime, we do not rule out the possibility that alternative configurations or optimization strategies could leverage its potential. Further research may uncover conditions under which vision transformer architectures can benefit more substantially from symmetric initialization.